Grade Retention and Mixed Race

### 2022-11-02 products/paper/main.qmd
Add rough draft of section on grade retention

-   Define what grade retention is and the era of "social promotion." Recent work suggests a decline in grade retention. Can cite Warren and Saliba on the trends.
-   Grade retention is consequential for later educational outcomes. CITE the evidence on this although, Warren and Saliba are also very iff-y on this. The Andrew article on "scarring" effects of grade retention is also a good one here as is the Hauser et. al. 2001 piece. Furthermore, little evidence on the positive effects of grade retention (again cite Hauser 2001). Grade retention also varies by race and socioeconomic background and is thus can be seen as a mechanism for the reproduction of class and racial inequality. In fact, because grade retention can happen quite early in the K-12 system, for many children, it may be the first direct encounter with a "sorting" mechanism for inequality.
-   There are many challenges to measuring grade retention in existing data sources. One of the most common techniques is to flag students who are behind the expected grade for their age. This technique is imperfect because
    -   it captures a cumulative prevalence measure of grade retention rather than incidence.
    -   it cannot distinguish cases in which children were delayed in beginning school ("redshirted") from grade retention.
    -   Students can reasonable be in two modal grades based on a given calendar year of age. Without detailed information on birthdates and survey timing, overall grade retention will be underestimated. For example, a child who is reported as 9 years old and in third grade could either be in the expected grade or could have been held back, depending on their birthday. A child who is 10 years old, however, is clearly behind expected grade.
-   Unfortunately, the only measure of grade retention available in the American Community Survey is based on students being clearly behind expected grade. The CPS has some advantages for better measuring grade retention, but lacks the sample size to precisely estimate parameters for biracial populations.
-   While being clearly behind expected grade is a problematic measure of grade retention for researchers seeking to better understand the process of grade retention itself, that is not my goal here. Rather, I am using this measure as an early and accessible measure of racial stratification for young children and adolescents. I can thus use it as a window to understand biracial students' placement within a racial hierarchy. This approach is similar to that employed by Rosenfeld (CITE) to examine the differential experiences of children from straight and LGBT families. From this perspective, the major potential problem is that the overall measure underestimates grade retention. However, because this underestimation is largely a function of birthdate, it is unlikely to bias the results presented here.
-   There is a question of how much racial inequality is present in grade retention rates. Hauser et. al. 2001 suggest that while Black and Latino students are more likely to have been held back, those differences are largely explained by differences in STUFF. However, other more recent research finds at least moderate racial differences, particularly between Black an White students, even after controlling for a variety of other variables (cite Warren and Sariba, Rosenfeld, others?). Therefore, understanding the effect of socioeconomic and other resources on the differences between monoracial groups is also key to understanding the larger research question, just as it is for biracial students.

>>>>

Grade retention is the practice of having students repeat a grade due to poor academic performance. The frequency of grade retention has fluctuated historically in response to changes in educational policy. Rates of grade retention increased throughout the late 20th century [@hauser_raceethnicity_2000] but have declined substantially since 2004 [@warren_patterns_2014].

Prior work on grade retention has failed to show a positive effect of grade retention, and has instead found evidence of "scarring" effects of grade retention that lead to later negative educational outcomes, such as a higher dropout risk [@jimerson_metaanalysis_2001; @stearns_staying_2007; @andrew_scarring_2014; @hughes_effect_2017]. The risk of grade retention also varies substantially by race and class background, with Black and Latino students at particularly high risk of grade retention relative to other groups [@warren_patterns_2014].

Because grade retention often happens very early in the K-12 system, it is one of the earliest mechanisms that individuals encounter for reproducing social inequality and stratification. For these same reasons, it can also provide an important window into understanding the placement of biracial children within a racially stratified educational environment.

Despite its long-standing use in the educational system, grade retention is famously difficult to measure accurately. In most data sources, researchers do not have access to specific reports of grade retention, but must rather infer grade retention from a discrepancy between a student's reported age and grade. Prior studies have used the concept of a student being behind modal grade [@bianchi_children_1984; @frederick_have_2008]. This measure captures the cumulative prevalence of grade retention, but not its incidence. It also cannot distinguish cases in which children were delayed in starting kindergartern from grade retention. The most notable limitation of this approach is the ambiguity created by the fact that at any given age, a student may reasonably be in two modal grades. Without detailed information on birthdates and survey timing, the correct modal grade for most students cannot be identified. This issue has been somewhat alleviated in prior work by the use of the October supplement to the Current Population Survey which is close to the beginning of the school year.

In the American Community Survey (ACS) data that I use here, I do not have the luxury of being able to identify or limit survey timing, so I instead use a related measure of whether a given student is clearly behind expected grade (CBEG). A student is considered CBEG if their age is higher than either of the expected ages for a student of that grade. This measure will underestimate overall grade retention because it will miss students who have been retained but not yet had a birthday in their current grade that would place them CBEG. However, the goal of this study is not to estimate grade retention accurately, but rather to understand racial differences in grade retention. Because this bias is largely a function of when students have birthdays, it should be more or less random with regard to sociodemographic characteristics. This approach is equivalent to the method used by @rosenfeld_nontraditional_2010 on 2000 Census data, but more detailed information on current grade in recent ACS data allows for more precise estimation.
